Linting Linearity in Core/System FC

### 2023-01-19 report.tex

\begin{figure}[h]
\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
    {\Gamma ; \Delta \vdash M \with N : A \with B}
    \qquad
    \infer*[right=($\with E_L$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
\qquad
    \infer*[right=($\with E_R$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Gamma ; \Delta \vdash M : A}
    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
\qquad
    \infer*[right=($\oplus I_R$)]
    {\Gamma ; \Delta \vdash M : B}
    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]
    \infer*[right=($\oplus E$)]
    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Gamma ; \Delta \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
\]
\caption{Typing rules for a linearly-typed lambda calculus}
\end{figure}

@@

Note that if the list is empty, we terminate the program with an error. This
function is unsafe -- our program might crash if we use it on an invalid input.
A common technique in Haskell is to use more expressive types to assert
properties about the values and get rid of the unsafe cases (e.g. a
\texttt{NonEmpty} is a list that can not be empty). A well liked motto is "make
invalid states unrepresentable". In this light, we will introduce Generalized
Algebraic Data Types (GADTs) and create a list type indexed by size for which we
can write a completely safe \texttt{head} function. This is done by expressing that the size
of the list must be at least one at the type level.

% TODO: But what are ADTs? Perhaps expand Haskell introduction section to
% include recursive ADTs
\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular feature with
which users can define data types as they would common recursive algebraic data
types, with the added abilitiy to give explicit type signatures for the data
constructors where the result type may differ in the type parameters across
branches. Pattern matching against GADTs can introduce local type refinements,
that is, pattern matching against certain data constructors will refine the type
information used for typechecking individual case alternatives. We develop
the length-indexed lists example without discussing the type system and type
inference details of GADTs which we later explore in~\ref{related-work-gadts}.

>>>>

\begin{figure}[h]
\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
    {\Gamma ; \Delta \vdash M \with N : A \with B}
    \qquad
    \infer*[right=($\with E_L$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
\qquad
    \infer*[right=($\with E_R$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Gamma ; \Delta \vdash M : A}
    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
\qquad
    \infer*[right=($\oplus I_R$)]
    {\Gamma ; \Delta \vdash M : B}
    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]
    \infer*[right=($\oplus E$)]
    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Gamma ; \cdot \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
\]
\caption{Typing rules for a linearly-typed lambda calculus}
\end{figure}

@@

If the list is empty, we terminate the program with an error, otherwise, we take
the first element. This function is unsafe -- our program might crash if we use
it on an invalid input.
\todo[inline]{When applied to the empty list, ...}

A common technique in Haskell is to use more expressive
types to assert properties about the values and get rid of the impossible cases
(e.g. a \textt{NonEmpty} is a list that can not be empty).

\todo[inline]{Leveraging Haskell's more advanced features, we can
  .... }

A well liked motto is
"make invalid states unrepresentable". In this light, we will introduce
Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
for which we can write a completely safe \texttt{head} function by expressing
that the size of the list must be at least one, at the type level.

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular\todo{nope} feature with
which users can define data types as they would common recursive algebraic data
types, with the added abilitiy to give explicit type signatures for the data
constructors where the result type may differ in the type parameters across
branches. Pattern matching against GADTs can introduce local type refinements,
that is, pattern matching against certain data constructors will refine the type
information used for typechecking individual case alternatives. We develop
the length-indexed lists example without discussing the type system and type
inference details of GADTs which we later explore in~\ref{related-work-gadts}.
